The natural history of scavenging in vertebrates

### 2016-03-11 natural history of scavenging.tex
Added new section
Aerial written and terrestrial started

\section*{Introduction}
%TG: it might be better to reorganise the intro as:
%§1 - scavenging is understudied because it's hard to say who's a scavenger
%§2 - it's even more problematic in the fossil record but people have tried some indirect approaches
%§3 - good thing is that we can apply these indirect approaches to extant systems
%§4 - in this review we discuss some stuff.
%TG: or something along these lines, obviously flexible!
Historically, scavengers have not been viewed as the most charismatic animals. This may go some way to explaining the gap in our knowledge of the prevalence of this behaviour. Prof Sanborn Tenney writing in 1877 for The American Naturalist had this to say about one well known group, “Prominent among the mammalian scavengers are the hyenas, the ugliest in their general appearance of all the flesh eaters.” He contrasts these with “nobler kinds” of carnivores such as lions and tigers. Even aside from our own subjective biases, scavenging is a difficult behaviour to detect after the fact. Without catching a carnivore in the act of killing we are left to infer how the prey was killed. Some simple heuristics can inform us, for instance, in cases where the prey item was simply too large to have been killed by the ostensible predator. But clearly, a scavenger doesn’t only feed on animals too big for it to have hunted. 
This is significant, because as we are now beginning to realise, the extent of this behaviour is enormous among the carnivores such that, “in some ecosystems, vertebrates have been documented to assimilate as much as 90\% of the available carrion" \citep{benbow2015introduction}. Even Tenney’s noble big cats are now known to take in a significant portion of carrion in their diet where some lion populations get over 50\% of their meat from carcasses.
The obvious lack of direct behavioural data compounds the difficulty of discerning scavenging among extinct forms. Indeed, a single species of dinosaur notwithstanding, a synthesis describing the natural history of scavengers is absent from the literature.

@@

\section{Aerial Scavengers}
Species capable of flight have effectively added an extra spatial dimension, i.e. the vertical component, to their sensory environment.
This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger.
Moreover, flight is a cheaper means of locomotion than running \citep{tucker1975energetic}. 
Thus, it appear that would-be scavengers have a distinct advantage by taking flight.

\subsection*{Cenozoic}
Birds are the dominant vertebrate fliers today and include the best known scavengers on Earth, the vultures. 
These birds consist of two convergent groups, old world and new world vultures and represent the only example of obligate vertebrate scavengers. 
They have a suite of adaptations that allow them to flourish in this niche.
Vultures extend the energetic advantage of flight further by engaging in soaring instead of flapping flight, which is even cheaper \citep{hedenstrom1993migration}.
Their efficiency is illustrated by cases of predators like bears and wolves benefiting by taking more carrion in their diet in areas bereft of vultures through competitive release \citep{devault2003scavenging}. 
In flight, birds possess a huge advantage over any terrestrial obligate scavenger. 
Flight affords them the ability to range over a much larger area and detect carrion from an elevated vantage point.
\cite{pennycuick1972soaring} conservatively estimated that a Gyps vulture could identify activity at a carcass 4 km away.

@@

\subsection*{Mesozoic}
In ancient ecosystems, the volant pterosaurs have also been postulated as existing in a vulture-like niche \citep{witton2008reappraisal}. 
Certain clades of these animals could reach enormous sizes (e.g. Azhdarchids) and look to have engaged in soaring flight. 
However, the inflexibility of their necks and straight, rather than hooked jaw morphology argues against their existing as obligate scavengers \citep{witton2008reappraisal}. 
As yet, no one has ever attempted an energetics approach for this group which is likely due to the many uncertainties over their biology \citep{witton2010size}.

\subsection*{Palaeozoic}
The absence of flying vertebrates in the Palaeozoic may have permitted terrestrial forms to take in a higher proportion of carrion in their diet.

>>>>

\section*{Introduction}
%TG: it might be better to reorganise the intro as:
%§1 - scavenging is understudied because it's hard to say who's a scavenger
%§2 - it's even more problematic in the fossil record but people have tried some indirect approaches
%§3 - good thing is that we can apply these indirect approaches to extant systems
%§4 - in this review we discuss some stuff.
%TG: or something along these lines, obviously flexible!
Historically, scavengers have not been viewed as the most charismatic of animals. This may go some way to explaining the gap in our knowledge of the prevalence of this behaviour. Consider Professor Sanborn Tenney writing in 1877 for The American Naturalist who had this to say about one well known group, “Prominent among the mammalian scavengers are the hyenas, the ugliest in their general appearance of all the flesh eaters.” He contrasts these with “nobler kinds” of carnivores such as lions and tigers. Even aside from our own subjective biases, scavenging is a difficult behaviour to detect after the fact. Without catching a carnivore in the act of killing we are left to infer how the prey was killed. Some simple heuristics can inform us, for instance, in cases where the prey item was simply too large to have been killed by the ostensible predator \citep{pobiner2008paleoecological}. But clearly, a scavenger doesn’t only feed on animals too big for it to have hunted. The obvious lack of direct behavioural data compounds the difficulty of discerning scavenging among extinct forms. Indeed, a single species of dinosaur notwithstanding, a synthesis describing the natural history of scavengers is absent from the literature. Fortunately, research in this area is on the rise. As a result we are now beginning to realise the extent of this behaviour such that, “in some ecosystems, vertebrates have been documented to assimilate as much as 90\% of the available carrion" \citep{benbow2015introduction}. Even Tenney’s noble big cats are now known to take in a significant portion of carrion in their diet where some lion populations get over 50\% of their meat from carcasses. A suite of methods have been used to discern the most suitable morphologies, physiologies and environments for a scavenging lifestyle to prosper. It is our aim in this review to employ these methods to gain an understanding of scavengers past and present.

The chief hurdle to scavenging is finding a sufficient quantity of food, the occurence of which is difficult to predict in space and time. The idea of scrounging from predator kills is undermined from studies showing that in the  majority of ecosystems more animals die from disease and starvation than predation \citep{benbow2015introduction}. Thus, any animal existing as a scavenger must maximise its detection capabilities and minimise its locomotory costs \citep{ruxton2004obligate}. The habitat must also be productive enough to sustain an animal biomass that will eventually produce carcasses.

\section{Aerial Scavengers}

Vultures represent the best known scavengers on Earth. These birds consist of two convergent groups, old world and new world vultures and represent the only example of obligate vertebrate scavengers today. Given their unique position, they have been extensively studied to determine what adaptations they possess that allows them to so flourish in this niche. As such, we can begin by exploring the adapdations and the environments of vultures to draw comparisons with other scavenging species and \textit{their} environments.

Species capable of flight have effectively added an extra spatial dimension, i.e. the vertical component, to their sensory environment over land animals. This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger. Such an ability has obvious benefits in detecting carrion. Vultures are known to have impressive visual acuity with one estimate indicating Lappet-faced Vultures (\textit{Torgos tracheliotus}) are capable of detecting a 2 metre carcass over 10 km away \citep{spiegel2013factors}. We know that many birds exist as facultative scavengers; storks, eagles, corvids, are all known to take substantial quantities of carrion in their diet. And eagles in particular are known to have highly developed visual abilities. It follows from this that the evolution of flight allowed aerial animals to detect far more carrion than their terrestrial counterparts.

Moreover, having a panoramic view means being able to gather a wealth of information from other foragers, be they conspecifics or other species. Again, returing to vultures, the genus \textit{Gyps} consists of highly social and colonially nesting species. These behaviours allow them forage far more efficiently because one bird can scrounge information on the location of food from another successful forager.

Flight is also cheaper means of locomotion than running \citep{tucker1975energetic}. This advantage can be extended further in larger species by engaging in soaring instead of flapping flight, which is even cheaper still (~ twice BMR) \citep{hedenstrom1993migration}. The advantages this confers are clear from the information we have on the enormous foraging ranges of some seabirds and accipiters. Indeed, it would be pointless to have incredible detection abilites and not have a cost efficient movement to benefit from it.

Avian flight originates in the Jurassic Period, conincident with the fossils of \textit{Archaeopteryx lithographica} so many of these benefits would have been realised from that point on for carnivorous birds. However, vertebrate flight is much older than this where pterosaurs predate bird origins by a considerable margin (the Triassic Period). Scavenging in this diverse group has been hypothesied many times before. Certain clades of these animals could reach enormous sizes (e.g. Azhdarchids with wingspans of 11 metres) and look to have engaged in soaring flight.  Although \cite{witton2008reappraisal} argued that neck inflexibility and straight, rather than hooked jaw morphology points against their existing as \textit{obligate} scavengers, Azhdarchid terrestrial proficency indicates they would have been comfortable foraging on the ground. Indeed, extant Marabou Storks have a comparable morphology and are noted facultative scavengers so it is reasonable to believe that certain pterosaurs behaved similarly.

Large body size confers substantial dominance benefits \citep{ruxton2004obligate}. Thus we would expect scavengers to have this trait selected for even in the case of weight-constrained fliers. Cinereous Vultures (\textit{Aegypius monachus}) and condors (\textit{Vultur gryphus}, \textit{Gymnogyps californianus}) all have body masses that can exceed 10 kg and represent some of the heaviest bird species capable of flight \citep{ferguson2001raptors,donazar2002effects}. And as we have noted the Azhdarchid pterosaurs were far bigger again, with estimated body masses of around 80 kg.

The only other vertebrate group capable of powered flight are the bats where scavenging has not been recorded to our knowledge. Their visual acuity is famously poor and echolocation does not lend itself to discovering immobile carrion. Their small size and poor terrestrial ability would also count against them at a carcass. The bat fossil record is notoriously poor owing to their fragile skeletons so we are unable to determine if extinct species were more suited to this lifestyle.

Vertebrate scavengers in general are responsible for the dispersal of nutrients. Consider the diversity of animals that can end up feeding at the carcass of an elephant. Here we have an incredibly dense and nutrient rich patch that ends up being distributed widely. Thus, in an ecological context the evolution of flight coupled with an ability of scavenge resulted in a world with a far more widely distributed nutrient landscape. In the absence of vertebrate scavengers, invertebrates and microorganisms would consume the carcass in-situ or at least distribute the constituent nutrients over a much shorter range.

\section{Terrestrial Scavengers}
Terrestrial scavengers can be thought of as existing in a two-dimensional plane while foraging for carrion directly. 
They can detect carcasses at a range that is defined by the radius of their sensory organs, usually the visual and olfactory senses. As a consequence they have a much more restricted view of the landscape than do aerial foragers. 
No contemporary terrestrial vertebrate exists as an obligate scavenger but most if not all carnivorous vertebrates are facultative scavenging behaviour to some extent. Terrstrial scavenging in the mammals is probably best known in an African context where hyenas, jackals and lions all take sizable proportions of carrion in their diet. In the case of the spotted hyena (\textit{Crocuta crocuta}) it can be as high as 99\% \citep{benbow2015introduction}. We can again use this species as our efficient terrestrial scavenger to compare with other species.

Similar to vultures they have well developed sensory organs, particularly in olfcation whereby they can detect a rotting carcass 2 km downwind. They have a characteristic "rocking horse gait"  which allows them to cover great distances. "Apart from being an important water-conservation strategy, nocturnal behavior may have evolved in the Hyaenidae as a means of reducing competition with the other dominant scavengers in African ecosystems, the vultures, which are exclusivly diurnal (Houston 1979)" Carnivore Behavior, Ecology, and Evolution By John L. Gittleman

@@

%\section{Aerial Scavengers}
Species capable of flight have effectively added an extra spatial dimension, i.e. the vertical component, to their sensory environment.
This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger.
Moreover, flight is a cheaper means of locomotion than running \citep{tucker1975energetic}. 
Thus, it appear that would-be scavengers have a distinct advantage by taking flight.

%\subsection*{Cenozoic}
Birds are the dominant vertebrate fliers today and include the best known scavengers on Earth, the vultures. 
These birds consist of two convergent groups, old world and new world vultures and represent the only example of obligate vertebrate scavengers. 
They have a suite of adaptations that allow them to flourish in this niche.
Vultures extend the energetic advantage of flight further by engaging in soaring instead of flapping flight, which is even cheaper \citep{hedenstrom1993migration}.
Their efficiency is illustrated by cases of predators like bears and wolves benefiting by taking more carrion in their diet in areas bereft of vultures through competitive release \citep{devault2003scavenging}. 
In flight, birds possess a huge advantage over any terrestrial obligate scavenger. 
Flight affords them the ability to range over a much larger area and detect carrion from an elevated vantage point.
\cite{pennycuick1972soaring} conservatively estimated that a Gyps vulture could identify activity at a carcass 4 km away.

@@

%\subsection*{Mesozoic}
In ancient ecosystems, the volant pterosaurs have also been postulated as existing in a vulture-like niche \citep{witton2008reappraisal}. 
Certain clades of these animals could reach enormous sizes (e.g. Azhdarchids) and look to have engaged in soaring flight. 
However, the inflexibility of their necks and straight, rather than hooked jaw morphology argues against their existing as obligate scavengers \citep{witton2008reappraisal}. 
As yet, no one has ever attempted an energetics approach for this group which is likely due to the many uncertainties over their biology \citep{witton2010size}.

%\subsection*{Palaeozoic}
The absence of flying vertebrates in the Palaeozoic may have permitted terrestrial forms to take in a higher proportion of carrion in their diet.
